---
title: "Between Multi-Test Functions Example"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}

multiBetween <- function(data){
    
    #vector to store p values
    p <- matrix(0, (nrow(data) - 1))
    
    #conduct all between subject anaovas
    for(i in 2:(nrow(data))){
        #conduct ANOVA
        a <- summary(aov(as.numeric(data[i, ]) ~ rep(c("control", "treatment"), each = 3)))
        #extract p value and update p values vector
        p[i - 1] <- as.numeric(unlist(a[[1]])["Pr(>F)1"])
    }
    
    #append p values to original dataset and add column names
    r <- data.frame(rbind(NA , p), data)
    colnames(r) <- c("ANOVA p", colnames(data))
    
    #return results: original dataset with the ANOVA p values appended
    return(r)
}


```

<br>

Before starting, remember to run all the code in the MultiBetweenFunctions.R file. This will make all the Paired Multi-Test functions available to you in R.

<br>

### Data Format

In your dataset, each row should represent a microbe and each column should represent a subject. The first row of the data should be a categorical variable describing the category the subject of that column belongs to. The values of the categorical variable should be characters (i.e., treatment/control rather than 0/1). 

Let's generate some simple data that fulfill these requirements and allow us to explore the Between Multi-Test functions:


```{r, eval = FALSE}
set.seed(1)       #set random number generator
d <- matrix(round(runif(60, 1, 100), 2), 10, 6)
cond <- rep(c("control", "treatment"), each = 3)
data <- rbind(cond, d)
rm(d, cond)
rownames(data)[2:11] <- paste("Microbe", letters[1:10])
colnames(data) <- paste("Subject", 1:6)
```


```{r, echo = FALSE}
set.seed(1)       #set random number generator
d <- matrix(round(runif(60, 1, 100), 2), 10, 6)
cond <- rep(c("control", "treatment"), each = 3)
data <- rbind.data.frame(cond, d)
rm(d, cond)
rownames(data)[2:11] <- paste("Microbe", letters[1:10])
colnames(data) <- paste("Subject", 1:6)
```

Let's look at the data the code above generated for us:

```{r}
data
```


As we can see, we have six subjects with three in the control group and three in the treatment group. For each subject we have the counts of 10 microbes.

<br>

### Conducting Between Subjects ANOVA

To conduct between subjects ANOVAs for each microbe, we can use the *multiBetween()* function:

```{r, eval = FALSE}
multiBetween(data)
```

```{r, echo = FALSE}
d <- cbind(multiBetween(data)[1:6], rep("...", 11))
colnames(d)[7] <- "..."
d
```

We can see that when we use this function, our data is returned to us with a column of between subject ANOVA *p* values appended to our data.

Notice that we don't have to specify which subjects are part of which groups, nor do we have to specify the number of groups we have. The function (really the *aov()* in *R*) will extract this information for us.

<br>

### Between Multi-Test Function Library

In this section, for each Paired Multi-Test function, I will list all possible arguments and possible values of the arguments including the argument's default.

<br>

Because R does not read spaces between arguments, rather than write the functions on a single line as we have done so far, like this:


```{r, eval = FALSE}
mean(x)
```


I will expand the function across multiple lines and comment the arguments like this:

```{r, eval = FALSE}
mean(
    x    #the numbers to find the mean of
    )
```

<br>

#### Between Multi-Test Functions:


#### multiBetween()

```{r , eval = FALSE}
multiBetween(
    data              #data as described at top of this document
)
```

<br>



#### Questions or comments

If you have any further questions, or comments, you can contact Matt Cook (cookm346@myumanitoba.ca).



<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>











